Uncertainty propagation for predictor effects and bias correction in generalized linear (mixed) models

### 2021-11-21 variable_predictions.Rnw
Trying to write

\title{Uncertainty propagation for variable (predictor) effects and bias correction in generalized linear (mixed) models???}

@@

In many applications, predictions of expected responses, or response probabilities, are often of interest. Predictions provide a great way to summarize what the regression model is telling us and are very useful for interpreting and visualizing models estimates using graphs. For example, in logistic regression models, the coefficient estimates are usually not easy to interpret, and we may want to explore the ``effects" of covariates on predicted probabilities, or predictions of responses at various levels of covariates -- \emph{predictor effects}. For example, as a way to plan, public health officials may want to know the ``effects" of household income, wealth index, etc, on the predicted probability of having improved water services among slum dwellers.

@@

When visually presented, predictor effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, a subject to less clarity of interpretation. Further, generating predictor effects together with the associated confidence intervals for regression models has a number of challenges. In particular: 
\begin{enumerate}
\item choice of representative values of \emph{focal} predictor(s) and appropriate ``model center" for \emph{non-focal} predictors especially in multivariate models
\item  propagation of uncertainty ([how] can we incorporate uncertainty in nonlinear components of (G)LMMs? Should we exclude variation in non-focal parameters?)
\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)M)
\end{enumerate}

@@

The outline of this article is as follows: Section \ref{sec:qoi} provides formal definition various quantities of interest, Section \ref{sec:stats_background} provides statistical background of estimation of model predictions, Section \ref{sec:uncertainity_propagation} describes mathematical and computation implementation of the proposed method for uncertainty propagation, and lastly Section \ref{sec:bias_correction} describes methods for bias correction in GL(M)M together with the computational implementation.

@@

Currently, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from the non-focal predictors when computing the CIs. A non-trivial way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal terms set to $0$ -- \emph{zeroing-out} variance-covariance matrix. This only works when the input predictors are \emph{centered} prior to model fitting, in case of numerical predictors, and even much complicated when the predictors are categorical. We first describe this variance-covariance based approach and then discuss our proposed method which is based on \emph{model centering} and does not require input predictors to be scaled prior to model fitting. Model center is found by averaging the variables over the model matrix, $\bX^\star$.

@@

The computation of $\hat{\eta}^\star$ remains the same as described above. However, to compute $\sigma$, $\Sigma$ is modified by \emph{zeroing-out} (the variance-covariance of all non-focal predictors are assigned zero) variances of non-focal terms. Although this is the simplest approach, it requires centering of continuous predictors prior to model fitting and proper way to average categorical predictors. One way to center predictors is to create a design matrix, $\bX$, and then compute $\bX_{c} = \bX - \bar{\bX}$.

@@

Consider centered model matrix $\bX^{\star}_{c} = \{\bX_f^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$. It follows that the non-focal terms in $\bX^{\star}_{c}$ are all zero. Consequently the uncertainty due to non-focal predictors are all zeroed-out in the computation of $\sigma$. More generally, centered design matrix, $\bX^{\star}_{c}$, impacts on the estimated value of the intercept and its associated variance but not the slopes. Thus since non-focal terms in $\bX^{\star}_{c}$ are all zero, it does not matter what their corresponding values are in the variance-covariance matrix. Hence, we can compute CIs from non-centered predictors (in other words, fitted models with predictors in their natural scales).

\subsubsection{Numeric interactions}

Consider a multiple linear regression model such that

\subsection{Simulation examples}

@@

\subsubsection{Continuous predictors}\label{sec:continuous_predictors}

@@

\subsubsection{Categorical predictors}\label{sec:categorical_predictors}

@@

\subsubsection{Linear mixed effect model}

@@

\subsubsection{Multiple outcomes}

>>>>

\title{Uncertainty propagation for predictor effects and bias correction in generalized linear (mixed) models}

@@

In many applications, predictions of expected responses, or response probabilities, are often of interest. Predictions provide a great way to summarize what the regression model is telling us and are very useful for interpreting and visualizing model estimates using graphs. For example, in logistic regression models, the coefficient estimates are usually not easy to interpret, and we may want to explore the ``effects" of covariates on predicted probabilities, or predictions of responses at various levels of covariates -- \emph{predictor effects}. For example, as a way to plan, public health officials may want to know the ``effects" of household income, wealth index, etc, on the predicted probability of having improved water services among slum dwellers.

@@

When visually presented, predictor effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, a subject to less clarity of interpretation. Further, generating predictor effects together with the associated confidence intervals for regression models has a number of challenges. In particular: 
\begin{enumerate}
\item choice of representative values of \emph{focal} predictor(s) and appropriate ``model center" for \emph{non-focal} predictors especially in multivariate models
\item  propagation of uncertainty ([how] can we incorporate uncertainty in nonlinear components of (G)L(M)Ms? Should we exclude variation in non-focal parameters?)
\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms)
\end{enumerate}

@@

The outline of this article is as follows: Section \ref{sec:qoi} provides formal definition various quantities of interest, Section \ref{sec:stats_background} provides statistical background of estimation of model predictions, Section \ref{sec:uncertainity_propagation} describes mathematical and computation implementation of the proposed method for uncertainty propagation, Section \ref{sec:bias_correction} describes methods for bias correction in GL(M)M together with the computational implementation, and lastly, in Section \ref{sec:simulation_example} we consider some simulation examples.

@@

Currently, commonly used \proglang{R} packages for constructing predictions do not exclude the uncertainties resulting from the non-focal predictors when computing the CIs. A non-trivial way to exclude uncertainties associated with non-focal predictors in some of these packages is to provide a user defined variance-covariance matrix with the covariances of non-focal terms set to $0$ -- \emph{zeroing-out} variance-covariance matrix. This only works when the input predictors are \emph{centered} prior to model fitting, in case of numerical predictors, and even much complicated when the predictors are categorical. We first describe this variance-covariance based approach and then discuss our proposed method which is based on \emph{model centering} and does not require input predictors to be scaled prior to model fitting.

@@

The computation of $\hat{\eta}^\star$ remains the same as described above. However, to compute $\sigma$, $\Sigma$ is modified by \emph{zeroing-out} (the variance-covariance of all non-focal predictors are set to zero) variances of non-focal terms. Although this is the simplest approach, it requires centering of continuous, i.e., $x_c = x - \bar{x}$, predictors prior to model fitting and proper way to average categorical predictors.

@@

Consider centered model matrix $\bX^{\star}_{c} = \{\bX_f^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$. It follows that the non-focal terms in $\bX^{\star}_{c}$ are all zero in simple models without interactions but are isolated (narrower) around the model center. Consequently the uncertainty due to non-focal predictors are isolated in the computation of $\sigma^2 = \textrm{Diag}(\bX^\star_c \Sigma \bX^{\star\top}_c)$. In addition, the computation of $\bX^{\star}_c$ impacts only on the intercepts and the non-focal terms, i.e., the slopes and variance of the focal predictors are not affected. This means that we can still generate isolated CIs without necessarily centering the predictors prior to model fitting.

In many applications, it usually important to report the estimates that reflect the expected values of the untransformed response. However, when dealing with nonlinear link functions, it is even harder to generate correct predictions that reflect the untransformed response due to the bias in the expected mean induced by the nonlinear transformation of the response variable. In such cases, bias correction is needed when back-transforming the predictions to the original scales. Most common approach for bias-adjustment is second-order Taylor approximation \citep{lenth2018package, duursma2003bias}. Here, we describe and implement a different approach, \emph{population-based} approach for bias correction.

\subsection{Population-based approach for bias correction}

The most precise (although not necessarily accurate!) way to predict is to condition on a value $F$ of the focal predictor and make predictions for all observations (members of the population) for which $x_f = F$ (or in a small range around $F$ ...). A key point is that the nonlinear transformation involved in these computations is always \emph{one-dimensional}; all of the multivariate computations required are at the stage of collapsing the multidimensional set of predictors for some subset of the population (e.g. all individuals with $x_f = F$) to a one-dimensional distribution of $\eta^\star$.

Once we have got our vector of $\eta^\star$ (which is essentially a set of samples from the distribution over $\eta^\star(x_f, \nset{x})$ for the conditional set), we may want a mean and confidence intervals on the mean on the response (data) scale, i.e. after back-transforming. In other words we can use the observations themselves then we just compute the individual values of $g^{-1}(\eta^\star)$ and compute their mean.

In population-based approach, $\eta^\star$ is constructed as a function of properly constituted levels of focal predictors, $x_f$, and entire population of the non-focal predictors, $\nset{x}$, as opposed to averaged non-focal predictors, $\nset{{\bar{x}}}$, described in the previous section. More specifically:
%
\begin{itemize}
\item compute linear predictor associated with the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
\item compute linear predictor associated with the focal predictors, $\eta_{jf} = \sum{\beta_f x_{jf}}$
\item for every value of the focal predictor, $\eta_{jf}$:
\begin{itemize}
\item $\eta_j^\star  = \eta_{jf} + \nset{\eta}$
\item $\hat{y}_j  = \textrm{mean} ~ g^{-1} \left(\eta_j^\star\right)$
\end{itemize}
\end{itemize}
%
We make similar adjustments to compute the variances of the predictions at every level of the focal predictor:
%
\begin{align}
\sigma_{jf}^2 = \textrm{Diag}(\bX^\star_{jc} \Sigma \bX^{\star\top}_{jc})
\end{align}
%
where $\bX^{\star}_{jc} = \{\bX_{jf}^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$ and 
%
\begin{align}
\mathrm{CI}_j = \mathrm{mean} ~ \eta_j^\star \pm q\sigma_{jf}
\end{align}
%

\section{Simulation examples}\label{sec:simulation_example}

@@

\subsection{Continuous predictors}\label{sec:continuous_predictors}

@@

\subsection{Categorical predictors}\label{sec:categorical_predictors}

@@

\subsection{Linear mixed effect model}

@@

\subsection{Multiple outcomes}
